---
title: "Class One by Dr. Pacifique NIZEYIMANA"
subtitle: "Class One"
author: "NISHYIRIMBERE Yves_101388"
date: "`r Sys.Date()`"
---


### <span style="color:#4682B4; font-weight:bold;">Important definitions in data science</span>

Statistics
Modelling
Computation
Calculus
### <span style="color:#4682B4; font-weight:bold;">Today’s class,we will study</span>


Installation of R
Installation of R studio
Calculator in r
Data importation in R
Data description using numerical measures and graphs.

#### <span style="color:#4682B4; font-weight:bold;">Data structure (Vectors, Factors, Lists, Data frame, matrix and arrays) Vector: The foundational R data structure</span>



Numeric


```{r}
x <- c(0.5, 0.6)
age<-c(20,35,32,29) 
summary(age)
```
```{r}
age<-c(35,24,18,24) 
mean(age)
```
 Logical
```{r}

x <- c(TRUE, FALSE) 
x <- c(T, F)

```
 Character
 
```{r}
x <- c("a","b","c" )
class<-c("M","F","M")
```
 
 Integer
```{r}
x <- 9:29
```
 Complex
```{r}
x <- c(1+0i, 2+4i)
```
Matrix and operation within matrices

```{r}
x<- matrix(1:6, nrow = 2, ncol = 3)
x
```
```{r}
(y<-matrix(1:6, nrow = 3, ncol = 2))
```
```{r}
x%*%y
```
List

```{r}
x <- list(1,"a", TRUE, 1 + 4i)
x
```
Factor

```{r}
x <- factor(c("yes","yes","no"))
```

Create a vector with NAs in it

```{r}
x <- c(1, 2, NA, 10, 3)
```

Return a logical vector indicating which elements are NA
```{r}
is.na(x)
```

```{r}
x <- c(1, 2, 4,"NA",5)
bad <- is.na(x)
print(bad)
```
```{r}
x[!bad]
## [1] "1" "2" "4" "NA" "5"
```
```{r}
#What if there are multiple R objects and you want to take the subset with no missing

x <- c(1, 2, NA, 4, NA, 5)
y <- c("a","b", NA,"d", NA,"f")
good <- complete.cases(x, y)
good
```
Coercion

If character is present, in a vector, R convert everything in the vector to character
strings.
If a vector only contains logical and numbers, R  will convert the logical to numbers,
Every true becomes a 1, and every FALSE becomes 0
```{r}
sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))
```
Create data Using data frame Data frame is more general than matrix How???
Because data frame can contain different modes of data (Numeric, character and so
on) Similar to what you can see in SPSS, SAS,.......


```{r}
### Let's create a  data frame

studentID<-c(1,2,3,4,5)
math_score<-c(12,17,10,9,NA)
gender<-c("M","F","M","M","F")
it_score<-c(13,18,11,10,19)
scoredata<-data.frame(studentID,gender,math_score,it_score)
scoredata
```
Create data from keyboard
Steps 1. Create data frame (or matrix) with variable names 2. Invoke the text editor in the data objected created at first step

```{r}
data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric(0),bmi=numeric(0))
data_class2<-edit(data_class2)

```


## Dat Importation

R has some features that can allow to import data from different sources ( It can be text file, spreadsheet, or database)
1.	Data from excel
2.	Data statistical packages (SAS, SPSS, Stata)
3.	Data from Text files (ASCII, XML, Webscraping)
4.	Data from database management systems (SQL,MySQL, Oracle, Access)

```{r}
# HW1: Import data from Statistical package and from Database management systems
data_class<-read.table("/Users/nishyirimbereyves/Downloads/class_data.txt",header = TRUE, sep = "\t") 
data_class
```
```{r}
head(data_class)
```
```{r}
tail(data_class)
```
```{r}
data_class[10:20,]
```
```{r}
summary(data_class$Height)

```
```{r}
length(data_class$Weight)
```
```{r}
data_class[,-1]
```

In case you want to use data set built in R

```{r}
data() # list of datasets currently available 

```

```{r}
data("airquality") variable.names(airquality)
```

```{r}
str(airquality)
```

Exploration data analysis

Examine numerical variable using common summary

```{r}
print(data_class)
```
```{r}
summary(data_class)
```
```{r}
str(data_class)
```

Summary Statistics

```{r}
summary(data_class)
```
```{r}
apply(data_class, 2,mean)
```

```{r}
 ## Height  Weight

apply(data_class,2,sd)
```

```{r}
c(mean(data_class$Height),sd(data_class$Height))
```

```{r}
c(mean(data_class$Weight),sd(data_class$Weight))
```

```{r}
c(Mean=mean(data_class$Height),SD=sd(data_class$Height))
```
```{r}
c(Mean=mean(data_class$Weight),SD=sd(data_class$Weight))
```

### Variation and covariation using both numerical and graphs
## Relationship between a categorical and a continuous variable

```{r}
data("iris")
head(iris)
```

```{r}
mean(iris$Petal.Length)
```

```{r}
mean(iris$Petal.Length[iris$Species=="setosa"])
```

```{r}
mean(iris$Petal.Length[iris$Species=="versicolor"])
```

```{r}
mean(iris$Petal.Length[iris$Species=="virginica"])
```
## Shortcut

```{r}
by(iris$Petal.Length,iris$Species,mean)
```
```{r}
by(iris$Petal.Length,iris$Species,sd)
```
```{r}
by(iris$Petal.Length,iris$Species,summary)
```
## Visualize the differences in continuous variables between categories using Box-and- whisker plot

```{r}
boxplot(iris$Sepal.Length~iris$Species,data=iris, main="Comparison")
```
### Relationship between two categorical variables

  
```{r}
ucba<-data.frame(UCBAdmissions)
head(ucba)
```
```{r}
cross<-xtabs(Freq~Gender+Admit,data=ucba)
(cross<-xtabs(Freq~Gender+Admit,data = ucba))
```
## Is there gender bias in UCB graduate admission process?

```{r}
cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
prop.table(cross2,1)
```
BarChart
```{r}
dat <- read.table(text ="ProdA ProdB ProdC ProdD
1 110 50 60 70
2 120 50 80 65", header= TRUE)

barplot(as.matrix(dat),beside=FALSE,col=c("Red","green"))
```
#barplot(as.matrix(dat),beside=TRUE,col=c("gold3","red"))

```{r}
dat <- read.table(text = "A B C D E F G
1 10 80 30 90 70 60 90
2 20 50 70 50 40 10 40
3 60 80 80 60 60 30 160
4 20 40 70 80 20 10 70", header = TRUE)
```

```{r}
barplot(as.matrix(dat))
```
Scatter Plot

```{r}
plot(Weight~Height,data=data_class)
```
```{r}
plot(data_class[,1],data_class[,2])
```
```{r}
cor(data_class$Height,data_class$Weight)
```
Linear Models
```{r}
plot(Weight~Height,data=data_class)
abline(lm(Weight~Height,data=data_class)$coefficient)
```
```{r}
data_lm<-lm(Weight~Height,data=data_class)
summary(data_lm)
```
```{r}
hist(data_class$Height)
```
```{r}
hist(data_class$Weight,col = "blue")
```

## Linear with other line

```{r}
plot(Weight~Height,data=data_class)
abline(lm(Weight~Height,data=data_class)$coefficient)
lines(lowess(data_class$Height,data_class$Weight),col="blue")
```

```{r}
cor(data_class$Height,data_class$Weight)
```


```{r}
model1<-lm(Weight~Height,data=data_class)
```

## Data management

## Create another Variable

```{r}
attach(data_class)
(BMI<-Weight/(Height/100)^2)
```
```{r}
(BMI<-round(Weight/(Height/100)^2,digit=2))
```

```{r}
head(cbind(data_class,BMI),n=5)
```
```{r}
new_data_class<-cbind(data_class,BMI)
tail(cbind(data_class,BMI),n=10)
 
```
Summary of BMI

```{r}
summary(BMI)
```

```{r}
stem(BMI,scale=2)
```


